Instructions for developing new pages for the MarineOmics webpage

### 2023-12-15 DRAFT_DNA_methylation.Rmd
meth - refes in downstream section

Thanks to advances in sample preparation and sequencing methods, there is an influx of research examining epigenomics in non-model systems. Broadly, epigenetics can be defined as changes to gene expression that do not arise from changes in the DNA sequence. DNA methylation, or addition of a methyl (CH<sub>3</sub>) group to a cytosine base adjacent to a guanine (CpG) is one of the more commonly studies epigenetic mechanisms, partly because of its environmental sensitivity and potential role in phenotypic plasticity. For a good review of epigenetic mechanisms including DNA methylation in marine organisms see @doi:10.1146/annurev-marine-010318-095114.

The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data to an existing genome, and discuss needs for potential downstream applications. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. @https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS). When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated. Since bisulfite treatment can damage DNA, newer library preparation and sequencing methods (ex. [NEBNext® Enzymatic Methyl-seq Kit; EM-seq](https://www.neb.com/en-us/products/e7120-nebnext-enzymatic-methyl-seq-kit#Protocols,%20Manuals%20&%20Usage) are starting to gain popularity. As EM-Seq produces loci-level data, alignment of EM-Seq data is similar to WGBS. Also, technologies like Nanopore and PacBio sequencing can directly detect DNA methylation without bisulfite conversion. As DNA passes through a nanopore or is sequenced by single-molecule real-time (SMRT) technology, changes in electrical current or fluorescence patterns can indicate the presence of methylated bases.

@@

For a good review of epigenetic mechanisms including DNA methylation in marine organisms see @doi:10.1146/annurev-marine-010318-095114. @https://doi.org/10.1111/1755-0998.13542 offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS). The tutorial below is based on WGBS, though the general workflow would be consistent.

@@

The tutorial below is based on WGBS from a species of marine intertebrate (*Montipora capitata* coral) examined in @https://doi.org/10.1111/1755-0998.13542, and all raw data can be accessed under NCBI Bioproject PRJNA691891 if you would like to follow the tutorial with this dataset. The general workflow would be consistent for other library preparation and sequencing methods that produce loci-level data. The tutorial indicates places where special analytical consideration should be given to other data types.

@@

The software used as part of this tutorial is **Bismark Bisulfite Mapper** [@10.1093/bioinformatics/btr167], which is one of the more widely-used softwares to align methylation data to a reference genome. Another commonly-used options include bwa-meth [@pedersen2014fast] and BSMAP [@10.1186/1471-2105-10-232] (see [@10.1093/bib/bbab021] and [@10.1016/j.csbj.2022.08.051] for comparisons of these software for plant and mammal species, respectively). As with any software it is best to be famililar with the [User Manual](https://felixkrueger.github.io/Bismark/).

Within Bismark, the tutorial uses Bowtie 2 [@10.1038/nmeth.1923] as the alignment software. Although the Bismark user manual covers the necessary Bowtie 2 parameters, it may be useful to refer to the [Bowtie 2 manual](https://bowtie-bio.sourceforge.net/bowtie2/manual.shtml).

@@

```{r meth-ratio, echo = FALSE, out.width = "70%", fig.align = "center", fig.cap = "Frequency distribution of methylation ratios for CpG dinucleotides in oyster gill tissue. @10.7717/peerj.215"}
knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
```

@@

-   Gene-level methylation analysis. It is worth considering the utility of converting loci-level resolution data to gene-level resolution data

-   Differential loci or region analyses. methylKit is a popular package to conduct this analysis

-   Feature overlap. Bedtools for quantitative assessment and IGV for visual assessment are valuable for this.

# References

>>>>

Thanks to advances in sample preparation and sequencing methods, there is an influx of research examining epigenomics in non-model systems. Broadly, epigenetics can be defined as changes to gene expression that do not arise from changes in the DNA sequence. DNA methylation, or addition of a methyl (CH<sub>3</sub>) group to a cytosine base adjacent to a guanine (CpG) is one of the more commonly studies epigenetic mechanisms, partly because of its environmental sensitivity and potential role in phenotypic plasticity. For a good review of epigenetic mechanisms including DNA methylation in marine organisms see [@doi:10.1146/annurev-marine-010318-095114](#refs).

The purpose of this tutorial is to provide an example of aligning bisulfite-treated and enzymatically converted DNA sequence data to an existing genome, and discuss needs for potential downstream applications. The principle behind creating bisulfite-treated DNA libraries is that when cytosines are not methylated, then they will be converted to uracil, which ultimately result in thymine nucleotides in the sequence data. [@https://doi.org/10.1111/1755-0998.13542](#refs) offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS). When aligning bisulfite-converted data to a reference genome the percent methylation at a given CpG loci is determined by examining the ratio of thymines to cytosines. For example if there is a given cytosine adjacent to a guanine (ie CpG locus) with 30% of the reads containing thymine, one would consider this cytosine loci (CpG) to be 70% methylated. Since bisulfite treatment can damage DNA, newer library preparation and sequencing methods (ex. [NEBNext® Enzymatic Methyl-seq Kit; EM-seq](https://www.neb.com/en-us/products/e7120-nebnext-enzymatic-methyl-seq-kit#Protocols,%20Manuals%20&%20Usage) are starting to gain popularity. As EM-Seq produces loci-level data, alignment of EM-Seq data is similar to WGBS. Also, technologies like Nanopore and PacBio sequencing can directly detect DNA methylation without bisulfite conversion. As DNA passes through a nanopore or is sequenced by single-molecule real-time (SMRT) technology, changes in electrical current or fluorescence patterns can indicate the presence of methylated bases.

@@

For a good review of epigenetic mechanisms including DNA methylation in marine organisms see [@doi:10.1146/annurev-marine-010318-095114](#refs). [@https://doi.org/10.1111/1755-0998.13542](#refs) offers a comparison of three methods for quantifying DNA methylation at single base-pair resolution using whole genome bisulfite sequencing (WGBS), reduced representation bisulfite sequencing (RRBS), and methyl-CpG binding domain bisulfite sequencing (MBDBS). The tutorial below is based on WGBS, though the general workflow would be consistent.

@@

The tutorial below is based on WGBS from a species of marine intertebrate (*Montipora capitata* coral) examined in [@https://doi.org/10.1111/1755-0998.13542](#refs), and all raw data can be accessed under NCBI Bioproject PRJNA691891 if you would like to follow the tutorial with this dataset. The general workflow would be consistent for other library preparation and sequencing methods that produce loci-level data. The tutorial indicates places where special analytical consideration should be given to other data types.

@@

The software used as part of this tutorial is **Bismark Bisulfite Mapper** [@10.1093/bioinformatics/btr167](#refs), which is one of the more widely-used softwares to align methylation data to a reference genome. Another commonly-used options include bwa-meth [@pedersen2014fast](#refs) and BSMAP [@10.1186/1471-2105-10-232](#refs) (see [@10.1093/bib/bbab021](#refs) and [@10.1016/j.csbj.2022.08.051] for comparisons of these software for plant and mammal species, respectively). As with any software it is best to be famililar with the [User Manual](https://felixkrueger.github.io/Bismark/).

Within Bismark, the tutorial uses Bowtie 2 [@10.1038/nmeth.1923](#refs) as the alignment software. Although the Bismark user manual covers the necessary Bowtie 2 parameters, it may be useful to refer to the [Bowtie 2 manual](https://bowtie-bio.sourceforge.net/bowtie2/manual.shtml).

@@

```{r meth-ratio, echo = FALSE, out.width = "70%", fig.align = "center", fig.cap = "Frequency distribution of methylation ratios for CpG dinucleotides in oyster gill tissue. [@10.7717/peerj.215](#refs)"}
knitr::include_graphics("DRAFT_DNA_methylation/meth_ratio.png")
```

@@

-   Integrating with population genetic data. example: [@10.1093/gbe/evad013](#refs)

-   Gene-level methylation analysis: see [@. It is worth considering the utility of converting loci-level resolution data to gene-level resolution data.

-   Differential loci methylation and/or region analyses. example: [@10.1186/s12864-022-08781-5](#refs). methylKit is a popular package to conduct this analysis differential methylation anlaysis
-   Feature overlap. see [@10.1186/s12864-022-08781-5](#refs) Bedtools for quantitative assessment and IGV for visual assessment are valuable for this.

# References {#refs}
